---
title: "Activity-based Data Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{activity_analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r setup, message=FALSE, warning=FALSE}
library(scMINER)
library(ggplot2)
library(dplyr)
```

# Introduction

The **driver activity estimation** is one of the most important features of scMINER. **Mathematically**, the activity of one driver is a type of mean of the expressions of its targets. And **biologically**, the activity can be interpreted as a measure that describes how actively the driver functions, like the enzymes in digesting their subtracts, kinase in activating their downstream genes. Given the gene expression profiles and networks, scMINER can estimate the activities of some predefined drivers, including not only transcription factors (TFs) but also signaling genes (SIGs). scMINER provides a few functions to effortlessly calculate the activities, identify the hidden drivers and visualize them in multiple ways.

In this tutorial, we will use PBMC14k dataset for demonstration purpose.

<details>
<summary>**How was the PBMC14K dataset generated?**</summary>
```{r, prepare-input-eset, echo=FALSE}
## Load the raw count matrix
data("pbmc14k_rawCount")

## Read the true labels
pbmc14k_trueLabel <- read.table(system.file("extdata/demo_pbmc14k/PBMC14k_trueLabel.txt.gz", package = "scMINER"), header = T, row.names = 1, sep = "\t", quote = "", stringsAsFactors = FALSE)

## The true_label much cover all cells in the expression matrix
table(colnames(pbmc14k_rawCount) %in% row.names(pbmc14k_trueLabel))

## Create the sparse eSet object using the true_label
pbmc14k_raw.eset <- createSparseEset(input_matrix = pbmc14k_rawCount, cellData = pbmc14k_trueLabel, featureData = NULL, projectID = "PBMC14k", addMetaData = TRUE)

## Filter he sparse eSet object using the auto mode
pbmc14k_filtered.eset <- filterSparseEset(pbmc14k_raw.eset, filter_mode = "auto", filter_type = "both")

## Normalize he sparse eSet object using the auto mode
pbmc14k_log2cpm.eset <- normalizeSparseEset(pbmc14k_filtered.eset, scale_factor = 1000000, log_base = 2, log_pseudoCount = 1)

## Add the clustering results from MICA
pbmc14k_log2cpm.eset <- addMICAoutput(pbmc14k_log2cpm.eset, mica_output_file = system.file("extdata/demo_pbmc14k/PBMC14k/MICA/clustering_UMAP_euclidean_20_2.05.txt", package = "scMINER"), visual_method = "umap")

## Add cell type annotation results
celltype_map <- c(`1`="CD4TN", `2`="CD4TCM", `3`="CD8TN", `4`="NK", `5`="B", `6`="Monocyte", `7`="CD4Treg")
pbmc14k_log2cpm.eset$cell_type <- as.character(celltype_map[pbmc14k_log2cpm.eset$clusterID])
```
</details>
\

# Calculate the activities

scMINER provides two functions, `getActivity_individual()` and `getActivity_inBatch()`, to calculate the driver activities:

## Calculate the activities per group

`getActivity_individual()` is designed to calculate the activities per group. It takes the network files as the input:
```{r calculate-activity-individually, eval=FALSE}
## let's use B cell as an example
activity_B.eset <- getActivity_individual(input_eset = pbmc14k_log2cpm.eset[, pData(pbmc14k_log2cpm.eset)$trueLabel == "B"],
                                          network_file.tf = system.file("extdata/demo_pbmc14k/PBMC14K/SJARACNe/B/TF/bt100_pc001/consensus_network_ncol_.txt", package = "scMINER"),
                                          network_file.sig = system.file("extdata/demo_pbmc14k/PBMC14K/SJARACNe/B/SIG/bt100_pc001/consensus_network_ncol_.txt", package = "scMINER"),
                                          driver_type = "TF_SIG")
```

If you need to calculate the activity for multiple groups, this is usually the case, you can do it using `getActivity_individual()` as shown above one by one and merge the esets after that. Or, scMINER privides another function, `getActivity_inBatch()`, to calculate the activity in batch:
```{r calculate-activity-in-batch}
## let's use B cell as an example
activity.eset <- getActivity_inBatch(input_eset = pbmc14k_log2cpm.eset, sjaracne_dir = system.file("extdata/demo_pbmc14k/PBMC14k/SJARACNe", package = "scMINER"),  group_name = "trueLabel", driver_type = "TF_SIG", activity_method = "mean", do.z_normalization = TRUE)
```

```{r save-activity-eset, eval=FALSE}
saveRDS(activity.eset, file = "/work-path/PBMC14k/DATA/activity.eset")
```


# Differential activity analysis

Similar to `getDE()`, scMINER provides a function, `getDA()`, to perform the differential activity analysis and identify the group-specific drivers.

```{r differential-activity-analysis-1}
## 1. To perform differential expression analysis in a 1-vs-rest manner for all groups
da_res1 <- getDA(input_eset = activity.eset, group_by = "cell_type", use_method = "t.test")
head(da_res1)
```

```{r differential-activity-analysis-2, eval=FALSE}
## 2. To perform differential expression analysis in a 1-vs-rest manner for one specific group
da_res2 <- getDA(input_eset = activity.eset, group_by = "cell_type", g1 = c("B"), use_method = "t.test")

## 3. To perform differential expression analysis in a rest-vs-1 manner for one specific group
da_res3 <- getDA(input_eset = activity.eset, group_by = "cell_type", g0 = c("B"), use_method = "t.test")

## 4. To perform differential expression analysis in a 1-vs-1 manner for any two groups
da_res4 <- getDA(input_eset = activity.eset, group_by = "cell_type", g1 = c("CD4Treg"), g0 = c("CD4TCM"), use_method = "t.test")
```

The `getTopFeatures()` function can aslo be used to easily extract the group-specific markers from the differential expression result:
```{r get-top-drivers}
top_drivers <- getTopFeatures(input_table = da_res1, number = 10, group_by = "g1_tag", sort_by = "log2FC", sort_decreasing = TRUE)

dim(top_drivers)

head(top_drivers)
```

**The activity-based analysis is done!**

<details>
<summary>**Session info**</summary>
```{r session-info}
sessioninfo::session_info()
```
</details>
\
